Two subtle problems with over-representation analysis

### 2024-07-02 manuscript/manuscript.Rmd
update

* Research integrity

@@

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset based on p-value.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running clusterProfiler.
After clusterProfiler using the parameters above, the number of gene sets in the results was
quantified, and compared to the number that should have been reported using the two gene
limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* values of 1 were appended to the p-values obtained by
clusterProfiler, followed by FDR correction in R, to obtain the properly corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.

@@

These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
Where it makes sense to use ORA would be when FCS is not applicable, for example when
investigating the over-represented categories in gene expression modules defined by packages
like WGCNA (Weighted Gene Correlation Network Analysis) [@Langfelder2008-hu], or when
investigating over-represented functional categories among genes with variants associated
with a disease.

>>>>

* Research rigour

@@

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset based on p-value.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running clusterProfiler.
After running clusterProfiler using the parameters above, the number of gene sets in the
results was quantified, and compared to the number that should have been reported using the
two gene limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* values of 1 were appended to the p-values obtained by
clusterProfiler, followed by FDR correction in R, to obtain the properly corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.

@@

These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that if the data can be scored/ranked, methods like FCS are recommended over ORA.
ORA is better suited to other cases where genes are classified into binary groups.
For example when investigating the over-represented categories in gene expression modules
[@Langfelder2008-hu], or when investigating over-represented functional categories among
genes with variants associated with a disease [@White2019-ps].
